A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-20 sections/introduction.tex

Modern functional programming languages, such as Haskell or OCaml,
use sophisticated forms of type inference. The type systems of these languages are
descendants of
Hindley-Milner~\cite{hindley1969principal,milner1978theory,damas1982principal}, 
which was revolutionary at the
time in allowing type-inference to proceed without any type
annotation. The traditional Hindley-Milner type system supports
top-level \emph{implicit (parametric) polymorphism}~\cite{reynolds1983types}. With implicit
polymorphism, type arguments of polymorphic functions are
automatically instantiated. Thus implicit polymorphism and the absense of
type annotations mean that the Hindley-Milner type system 
strikes a great balance between expressive power and usability.

@@

Although type inference is important in practice and also in
academic research, there is little work in mechanically formalizing
such advanced forms of type inference in theorem provers.
The remarkable exception is some work done on the formalization of 
certain parts of Hindley-Milner type inference~\cite{naraschewski1999type,
dubois2000proving,dubois1999certification,urban2008nominal,
garrigue2015certified}. However
there is still no formalization of the higher-ranked type systems
that are employed by modern languages like Haskell.
This is at
odds with the current trend of mechanical formalizations in
programming language research. In particular both the POPLMark
challenge~\cite{aydemir2005mechanized} and
CompCert ~\cite{leroy2012compcert} have significantly promoted
the use of theorem provers to model various aspects of programming
languages. Today papers in various programming language venues routinely
use theorem provers to mechanically formalize: \emph{dynamic and
  static semantics} and their correctness properties~\cite{aydemir2008engineering},
\emph{compiler correctness}~\cite{leroy2012compcert}, \emph{correctness of
  optimizations}~\cite{Bertot04}, \emph{program analisys}~\cite{Chang2006}
or proofs involving \emph{logical relations}~\cite{abel2018}. The
main argument for mechanical formalizations is a simple one. Proofs
for programming languages tend to be \emph{long}, \emph{tedious} and
\emph{error-prone}. In such proofs it is very easy to make mistakes
that may invalidate the whole development. Furthermore, readers and
reviewers often do not have time to look at the proofs carefully to
check their correctness. Therefore errors can go unnoticed for a
long time.  Mechanical formalizations provide, in principle, a natural
solution for these problems. Theorem provers can automatically check and
validate the proofs, which removes the burden from checking from both
the person doing the proofs as well as readers or reviewers.

@@

Due to those two challenges, our work takes a different approach by refining and
extending the idea of \emph{worklists judgments}~\cite{itp2018},
proposed recently to mechanically formalize an algorithm for
\emph{polymorphic subtyping}. A key innovation in our work is how
to adapt the idea or worklist judgments with
\emph{inference judgements}, which are not needed for polymorphic subtyping.
In our work we show how worklist
judgements can be extended to deal with the inference judgements that
are necessary for type-inference.  The idea is to use a continuation
passing style to enable the transfer of inferred information across
judgements. A further refinement to the idea of worklist judgements is
the \emph{unification between ordered
  contexts~\cite{gundry2010type,dunfield2013complete} and worklists}.  This
enables precise scope tracking of free variables in
judgements. Furthermore it avoids the duplication of context
information across judgments in worklists that occurs in other
techniques~\cite{Reed2009,Abel2011higher}.
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. The calculus and its metatheory
have been fully formalized in the Abella theorem prover~\cite{AbellaDesc}.

@@

\item {\bf Unified judgment worklists and contexts:} Another technical contribution is the unification
  between ordered contexts and judgement worklists. This enables precise scope tracking
  of variables in judgements, and avoids the duplication of context information across
  judgments in worklists.

>>>>

Modern functional programming languages, such as Haskell or OCaml,
use sophisticated forms of type inference. The type systems of these languages are
descendants of
Hindley-Milner~\cite{hindley1969principal,milner1978theory,damas1982principal}, 
which was revolutionary at the
time in allowing type-inference to proceed without any type
annotation. The traditional Hindley-Milner type system supports
top-level \emph{implicit (parametric) polymorphism}~\cite{reynolds1983types}. With implicit
polymorphism, type arguments of polymorphic functions are
automatically instantiated. Thus implicit polymorphism and the absence of
type annotations mean that the Hindley-Milner type system 
strikes a great balance between expressive power and usability.

@@

Although type inference is important in practice and receives a lot of
attention in
academic research, there is little work on mechanically formalizing
such advanced forms of type inference in theorem provers.
The remarkable exception is some work done on the formalization of 
certain parts of Hindley-Milner type inference~\cite{naraschewski1999type,
dubois2000proving,dubois1999certification,urban2008nominal,
garrigue2015certified}. However
there is still no formalization of the higher-ranked type systems
that are employed by modern languages like Haskell.
This is at
odds with the current trend of mechanical formalizations in
programming language research. In particular both the POPLMark
challenge~\cite{aydemir2005mechanized} and
CompCert ~\cite{leroy2012compcert} have significantly promoted
the use of theorem provers to model various aspects of programming
languages. Today papers in various programming language venues routinely
use theorem provers to mechanically formalize: \emph{dynamic and
  static semantics} and their correctness properties~\cite{aydemir2008engineering},
\emph{compiler correctness}~\cite{leroy2012compcert}, \emph{correctness of
  optimizations}~\cite{Bertot04}, \emph{program analisys}~\cite{Chang2006}
or proofs involving \emph{logical relations}~\cite{abel2018}. The
main argument for mechanical formalizations is a simple one. Proofs
for programming languages tend to be \emph{long}, \emph{tedious} and
\emph{error-prone}. In such proofs it is very easy to make mistakes
that may invalidate the whole development. Furthermore, readers and
reviewers often do not have time to look at the proofs carefully to
check their correctness. Therefore errors can go unnoticed for a
long time.  Mechanical formalizations provide, in principle, a natural
solution for these problems. Theorem provers can automatically check and
validate the proofs, which removes the burden from checking from both
the person doing the proofs as well as readers or reviewers.

@@

Due to those two challenges, our work takes a different approach by refining and
extending the idea of \emph{worklists judgments}~\cite{itp2018},
proposed recently to mechanically formalize an algorithm for
\emph{polymorphic subtyping}. A key innovation in our work is how
to adapt the idea of worklist judgments to
\emph{inference judgements}, which are not needed for polymorphic subtyping.
In our work we show how worklist
judgements can be extended to deal with the inference judgements that
are necessary for type-inference.  The idea is to use a continuation
passing style to enable the transfer of inferred information across
judgements. A further refinement to the idea of worklist judgements is
the \emph{unification between ordered
  contexts~\cite{gundry2010type,dunfield2013complete} and worklists}.  This
enables precise scope tracking of free variables in
judgements. Furthermore it avoids the duplication of context
information across judgments in worklists that occurs in other
techniques~\cite{Reed2009,Abel2011higher}.
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. The calculus and its metatheory
have been fully formalized in the Abella theorem prover~\cite{AbellaDesc}.

@@

\item {\bf Unification of judgment worklists and contexts:} Another technical contribution is the unification
  between ordered contexts and judgement worklists. This enables precise scope tracking
  of variables in judgements, and avoids the duplication of context information across
  judgments in worklists.
